\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-05 honors_thesis/journal_paper/paper.tex
Finished revisions of the journal paper.  It is almost ready to submit

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{\vec{V}}}
\providecommand{\specpos}{\ensuremath{\vec{X}}}
\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casegen}{\ensuremath{c}}
\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
\providecommand{\casexn}{\ensuremath{(S,-)}}
\providecommand{\casexx}{\ensuremath{(S,S)}}
\providecommand{\casexl}{\ensuremath{(S,\leftind)}}
\providecommand{\casexr}{\ensuremath{(S,\rightind)}}
\providecommand{\casepn}{\ensuremath{(-,-)}}
\providecommand{\casepl}{\ensuremath{(-,\leftind)}}
\providecommand{\casepr}{\ensuremath{(-,\rightind)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

To solve this problem, we propose a method for performing two iterations of PSO
at the same in parallel that we call speculative evaluation.  The name comes
from an analogy to speculative execution techniques commonly used in
processors.  Modern processors, when faced with a branch on which they must
wait (e.g. a memory cache miss), guess which way the branch will go and start
executing, being careful not to make changes that can not be undone.  If the
processor guesses right, execution is much farther ahead than if it had just
idly waited on the memory reference.  If it guesses wrong, execution restarts
where it would have been anyway.  This is called speculative execution or
branch prediction.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one (or
more) neighbors to either side of it; in the Complete topology, every particle
is a neighbor to every other particle.  It is generally assumed that, a
particle is also a neighbor to itself in that its own position and value is
considered when updating the particle's personal best, $\nbest$.  Thus with $p$
particles, using the Ring topology each particle has three neighbors: $i-1$,
$i$ (itself), and $i+1$.  With the Complete topology, each particle has $p$
neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to simulate the behavior
of Complete with far less communication~\cite{mcnabb-cec09}.  In the Random
topology, each particle randomly picks two other particles to share information
with at each iteration, along with itself.  Thus in both the Ring and the
Random topologies, all particles have 3 neighbors.

@@

To address this issue we present here a different use of additional processors.
Instead of adding particles we employ a speculative evaluation approach as
described in the introduction.  This allows us to perform iterations two at a
time.

@@

Our refactoring hinges on the idea that there is a finite set of possible
updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
are discussing, there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update; a hyphen represents no update, $L$ represents an update
to $\nbest$ coming from the left neighbor, $R$ represents an update to $\nbest$
coming from the right neighbor, and $S$ represents an update to either $\pbest$
or $\nbest$ coming from the particle itself.  In the equations that follow, we
refer to an unspecified update case as $\casegen$, and to the set of cases
collectively as $\caseset$

@@

The value of the refactoring can best be seen in a small example.  Suppose that
unlimited processors were available and that the evaluation of the objective
function took 1 hour. Suppose also that for each particle the computations
$\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$ were sent to
separate processors and completed in parallel.  Once the values of
$\ofunc(\ppos_{t})$ have been computed for all particles, the computation
needed to determine the values of the indicator functions $\indic_{t+1}^{c}$
and to complete iteration $t+1$ is inconsequential.  In this environment all of
the function evaluations needed for 2 iterations of PSO could be completed in
just one hour where as the conventional implementation would require two hours
for those same 2 iterations.

@@

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration, where
$p$ is the swarm size.  Clearly it is much more desirable to have a sparse
topology, where $O(np)$ is much smaller than $O(p^2)$.  However, some functions
are better optimized with the Complete topology and the quick spread of
information it entails than with sparse topologies.  Accordingly, we use the
Random topology described in~\cite{mcnabb-cec09}, which has been shown to
effectively simulate the Complete topology.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology and
the Random topology on a number of common benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  First we discuss the
relatively easy case of a centralized parallel PSO algorithm with a master
computer and many slaves.  In such an architecture, the master keeps track of
all necessary information with only trivial message passing needed.  Then we
discuss the more complicated case of a distributed algorithm, where each
particle is on its own and needs to send and receive messages to and from other
particles.  Finally we discuss the further complications of a dynamic topology,
where a particle's neighbors change from one iteration to another.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
an architecture, the master generates the positions to evaluate speculatively
and decides which one to accept for each particle.  The outline of the
procedure is given in \alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
Table~\ref{tab:evals}).  Each of those combinations defines a speculative
particle, $\s_t$, that can be moved using the motion equations to obtain
$\noeval{\s}_{t+1}$.  The result is a set of particles $\noeval{\pset}_t$, and
for each particle a set of speculative children $\noeval{\sset}_{t+1}$, which
can all be evaluated at once.

@@

This is done by first updating each particle $\noeval{\p}_t$ to
$\nonbest{\p}_t$ and each speculative particle $\noeval{\s}_{t+1}$ to
$\nonbest{\s}_{t+1}$ by associating the value determined by the slave to the
particle it came from, and updating that particle's personal best if the value
reported is better than its previous personal best value.  Then each
$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
position and value with information about each of its neighbors.  This is
simply the original PSO algorithm, and corresponds to steps 1--5 in
\alg{centralized}.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected (step 7 in \alg{centralized}).

@@

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion of speculative
evaluations.  If speculative evaluation were not being performed, it is
possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
parallel PSO, so it is still quite possible that Random would be used even with
standard PSO.  But, to be fair in our comparisons, we compare to the standard
PSO algorithm using both the Random topology with two neighbors and the
Complete topology.  All functions have 20 dimensions, and all results shown are
averages over 20 runs.  We use 240 processors in our experiments (except where
otherwise noted), so our speculative algorithm uses 30 particles and standard
PSO uses 240 particles.  For ease of labeling, we call our speculative
algorithm SEPSO and standard PSO just PSO, with the topology following.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum.  Griewank has a global optimum with a value of 0, and
sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
showing average function value at each iteration, a more accurate plot for
Griewank shows the percent of runs that have found the global optimum by each
iteration.  We show results in \fig{basic-griewank2} for swarms of size 30 and
240 using the Ring topology.  One can see in the figure that SEPSO algorithm
finds the optimum faster than the original PSO.  However, because the swarm
size is so small, SEPSO gets stuck a little less than half of the time.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches that branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The branch that PSO would have taken is ignored, but $\p_t$ must still be
created so it can give the right personal best and neighborhood best values and
positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.

@@

In Table~\ref{tab:evals} we presented all possible branches that a particle
with two neighbors could take.  Here we lump all of the neighbors together and
consider the statistics for the five branches shown in Table~\ref{tab:branches}.

\begin{table}[ht]
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neighbor\\
  \end{tabular}
\end{table}

\begin{table}[ht]
  \caption{Branch Statistics in PSO}
  \label{tab:stats}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
	Topology&Function&1&2&3&4&5\\
	\hline
	\hline
	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
	\cline{2-7}
	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
	{7.6\%}\\
	\hline
	\hline
	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
	\cline{2-7}
	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
	{2.2\%}\\
	\hline
	\hline
	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
	\cline{2-7}
	&Average&{37.6\%}&{6.9\%}&{.2\%}&{45.5\%}&
	{9.8\%}\\
	\hline
	\multicolumn{7}{l}{The numbers in the column headings refer to branch
	numbers in Table~\ref{tab:branches}.}
  \end{tabular}
\end{table}

Row 1 in Table~\ref{tab:branches} corresponds to the phenomenon commonly called
stagnation in the literature, but only with a single particle.  As an
interesting aside, we found that complete stagnation (where the entire swarm is
stagnant) practically never occurs in PSO, though frequently large percentages
of the swarm are stagnant.  Iterations where the entire swarm is stagnant
average less than 2 in 1000 iterations, unless the swarm converges beyond
machine precision.  When graphs of the best function value flat-line, what
really has happened is a contraction of the particles' velocity, not stagnation
in the technical sense.

We seek to find the probability of taking any given branch, given whatever
information is needed: $\prob(B_t|\cdot)$.  In finding these probabilities, we
do not attempt to derive any distribution from the PSO equations, we simply
look at empirical distributions.  However, even with empirical distributions,
the problem with this approach is that it is not clear what information
influences the probability of taking a branch.  The failure of the naive
approach seems to say that the previous branch, $B_{t+1}$, is not incredibly
useful.  We look at two factors that we believe have a significant influence on
$\prob(B_t)$: topology and function.  Thus we are looking at $\prob(B_t|T,F)$.

>>>>

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{\vec{V}}}
\providecommand{\specpos}{\ensuremath{\vec{X}}}
\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casegen}{\ensuremath{c}}
\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
\providecommand{\casexn}{\ensuremath{(S,-)}}
\providecommand{\casexx}{\ensuremath{(S,S)}}
\providecommand{\casexl}{\ensuremath{(S,\leftind)}}
\providecommand{\casexr}{\ensuremath{(S,\rightind)}}
\providecommand{\casepn}{\ensuremath{(-,-)}}
\providecommand{\casepl}{\ensuremath{(-,\leftind)}}
\providecommand{\casepr}{\ensuremath{(-,\rightind)}}
\providecommand{\casepN}{\ensuremath{(-,N)}}
\providecommand{\casexN}{\ensuremath{(S,N)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) techniques commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, being careful not to make
changes that can not be undone.  If the processor guesses right, execution is
much farther ahead than if it had just idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies a particle is also a
neighbor to itself in that its own position and value are considered when
updating the particle's neighborhood best, $\nbest$.  Thus with $p$ particles,
using the Ring topology each particle with index $i$ has three neighbors:
$i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each particle has
$p$ neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to approximate the
behavior of Complete with far less communication~\cite{mcnabb-cec09}.  In the
Random topology, each particle randomly picks two other particles to share
information with at each iteration, along with itself.  Thus in both the Ring
and the Random topologies, all particles have three neighbors.

@@

To address this issue we present here a different use of additional processors.
Instead of adding particles we employ a speculative evaluation approach that
allows us to perform iterations two at a time.

@@

Our refactoring hinges on the idea that there is a finite set of possible
updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
are discussing there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update; a hyphen represents no update, $L$ represents an update
to $\nbest$ coming from the left neighbor, $R$ represents an update to $\nbest$
coming from the right neighbor, and $S$ represents an update to either $\pbest$
or $\nbest$ coming from the particle itself.  As an example, $\casexn$ refers
to the case that the particle finds a new personal best, but neither it nor its
neighbors found a position that updated its neighborhood best.  In the
equations that follow, we refer to an unspecified update case as $\casegen$,
and to the set of cases collectively as $\caseset$

@@

The value of the refactoring can best be seen in a small example.  Suppose that
unlimited processors were available and that the evaluation of the objective
function took one hour. Suppose also that for each particle the computations
$\ofunc(\ppos_{t})$ and all of the $\ofunc(\specpos_{t+1}^{c})$ were sent to
separate processors and completed in parallel.  Once the values of
$\ofunc(\ppos_{t})$ have been computed for all particles, the computation
needed to determine the values of the indicator functions $\indic_{t+1}^{c}$
and to complete iteration $t+1$ is inconsequential.  In this environment all of
the function evaluations needed for two iterations of PSO could be completed in
just one hour where as the conventional implementation would require two hours
for those same two iterations.

@@

The number of speculative evaluations needed per particle depends on the number
of neighbors each particle has.  The number of update cases in a topology where
each particle has $n$ neighbors is $2(n+1)$; there are two possibilities for
updates to $\pbest$ (updated by the particle itself and not updated), and $n+1$
possibilities for updates to $\nbest$ (updated by each neighboring particle and
not updated).  When the particle is also a neighbor to itself, as is always the
case in commonly used topologies, one of the cases can be eliminated, as a
particle cannot be the source of an update to its neighborhood best while not
also updating its personal best.  Thus we have $2(n+1)-1$, or $2n+1$,
speculative evaluations per particle.  In a swarm with $p$ particles and $n$
neighbors per particle, $(2n+1)p$ speculative evaluations are needed.

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In~\sec{results} we report the results for performing
speculative evaluation using both the Ring topology and the Random topology on
a number of common benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  In the rest of this
section we move from the mathematics of our method to its implementation.
First we discuss the relatively easy case of a centralized parallel PSO
algorithm with a master computer and many slaves.  In such an architecture, the
master keeps track of all necessary information with only trivial message
passing needed.  Then we discuss the more complicated case of a distributed
algorithm, where each particle is on its own and needs to send and receive
messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
an architecture, the master generates the positions to evaluate speculatively
as in Equation~\eqref{eq:pos2update}.  After having the slaves evaluate the
objective function at all necessary positions, the master then decides which
position to accept for each particle, as in Equation~\eqref{eq:val2updatelong}.
The outline of the procedure is given in \alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  Equations~\eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$ according to
	  Equation~\eqref{eq:pos2update}.
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$ according to
	  Equation~\eqref{eq:val2updatelong}.
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
Table~\ref{tab:evals}).  Each of those combinations defines a speculative
particle, $\s_t$, that can be moved using Equation~\eqref{eq:pos2update} to
obtain $\noeval{\s}_{t+1}$.  The result is a set of particles
$\noeval{\pset}_t$, and for each particle a set of speculative children
$\noeval{\sset}_{t+1}$, which can all be evaluated at once.

@@

This is done by first updating each particle's personal and neighborhood bests.
Mathematically, this corresponds to the evaluation of an indicator function
similar to that found in Equation~\eqref{eq:deficasexn}.  In practice, this is
done by moving each $\noeval{\p}_t$ to $\nonbest{\p}_t$ and each speculative
particle $\noeval{\s}_{t+1}$ to $\nonbest{\s}_{t+1}$ by associating the value
determined by the slave to the particle it came from, and updating that
particle's personal best if the value reported is better than its previous
personal best value.  Then each $\nonbest{\p}_t$ must be updated to $\p_t$ by
updating its neighborhood best position and value with information about each
of its neighbors.  This is simply the original PSO algorithm, and corresponds
to steps 1--5 in \alg{centralized}.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected as per Equation~\eqref{eq:val2updatelong} (step 7 in
\alg{centralized}).

@@

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion in the number of
speculative evaluations.  If speculative evaluation were not being performed,
it is possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
parallel PSO, so it is still quite possible that Random would be used even with
standard PSO.  But, to be fair in our comparisons, we compare to the standard
PSO algorithm using both the Random topology with two neighbors and the
Complete topology.  All functions have 20 dimensions, and all results shown are
averages over 20 runs.  We use 240 processors in our experiments (except where
otherwise noted), so our speculative algorithm uses 30 particles and standard
PSO uses 240 particles.  For ease of labeling, we call our speculative
algorithm SEPSO and standard PSO just PSO, with the topology following.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum.  Griewank has a global optimum with a value of 0, and
sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
showing average function value at each iteration, a more accurate plot for
Griewank shows the percent of runs that have found the global optimum by each
iteration.  We show results in \fig{basic-griewank2} for swarms of size 30 and
240 using the Ring topology.  One can see in the figure that the SEPSO
algorithm finds the optimum faster than the original PSO.  However, because the
swarm size is so small, SEPSO gets stuck a little less than half of the time.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches the branch that the original PSO would have taken (the
``behavior preserving'' branch), we take the child that has the best value.
The methodology is exactly the same as above except for the process of choosing
which speculative child to accept.  The only change needed in \alg{centralized}
is in step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.

@@

In Table~\ref{tab:evals} we presented all possible branches that a particle
with two neighbors could take.  Here we lump all of the neighbors together and
consider the statistics for the five branches shown in
Table~\ref{tab:branches}.  In the identifiers, N represents an update to
$\nbest$ coming from any neighbor.

\begin{table}[ht]
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{lcc}
	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
	\hline
	\hline
	$\casepn$&No update&No update\\
	\hline
	$\casexn$&Self&No update\\
	\hline
	$\casexx$&Self&Self\\
	\hline
	$\casepl$&No update&Some Neighbor\\
	\hline
	$\casexl$&Self&Some Neighbor\\
	\hline
  \end{tabular}
\end{table}

\begin{table}[ht]
  \caption{Branch Statistics in PSO}
  \label{tab:stats}
  \centering
  \scriptsize
  \begin{tabular}{c|c|c|c|c|c|c}
	Topology&Function&\casepn&\casexn&\casexx&\casepN&\casexN\\
	\hline
	\hline
	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
	\cline{2-7}
	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
	{7.6\%}\\
	\hline
	\hline
	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
	\cline{2-7}
	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
	{2.2\%}\\
	\hline
	\hline
	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
	\cline{2-7}
	&Average&{37.6\%}&{6.9\%}&{0.2\%}&{45.5\%}&
	{9.8\%}\\
	\hline
  \end{tabular}
\end{table}

Row 1 in Table~\ref{tab:branches}, \casepn, corresponds to the phenomenon
commonly called stagnation in the literature, but only with a single particle.
As an interesting aside, we found that complete stagnation (where the entire
swarm is stagnant) practically never occurs in PSO, though frequently large
percentages of the swarm are stagnant.  Iterations where the entire swarm is
stagnant average less than 2 in 1000 iterations, unless the swarm converges
beyond machine precision.  When graphs of the best function value flat-line,
what really has happened is a contraction of the particles' velocity, not
stagnation in the technical sense.

We seek to find the probability of taking any given branch, given whatever
information is needed: $\prob(\caseset_t|\cdot)$.  In finding these
probabilities, we do not attempt to derive any distribution from the PSO
equations, we simply look at empirical distributions.  However, even with
empirical distributions, the problem with this approach is that it is not clear
what information influences the probability of taking a branch.  The failure of
the naive approach seems to say that the previous branch, $\caseset_{t-1}$, is
not incredibly useful.  We look at two factors that we believe have a
significant influence on $\prob(\caseset_t)$: topology and function.  Thus we
are looking at $\prob(\caseset_t|T,F)$.

We show in Table~\ref{tab:stats} with what percentage a particle takes each of
these branches for three different topologies and four different functions.
All of our statistics are from swarms of 240 particles.  Brief experimentation
showed that other swarm sizes had similar statistics.  We ran 750 iterations on
all combinations of functions and topologies except for the functions Griewank
and Rastrigin with the Complete topology.  We found that those runs frequently
converged past machine precision after 500 iterations, and that led to
erroneously high values for the probability of stagnation.  Instead we ran for
only 450 iterations on those two combinations.  All of our results were
averaged over 20 runs of the algorithm; thus the probabilities presented are
the averages of 3.6 million trials for the branch taken (2.16 million for the
two with only 450 iterations).  Table~\ref{tab:stats} contains the results.  We
have already defined the functions Sphere, Griewank, and Rastrigin.  Rosenbrock
is defined as $f(\Vec{x}) = \sum_{i=1}^D
\left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.
